Bootstrap learning via modular concept discovery

### 2013-04-22 drafts/ijcai2013/ijcai13.tex
More edits to paper and added missing citation.

\maketitle
\begin{abstract}
%\small
 Suppose a learner is faced with a domain of problems about which it
 knows nearly nothing. It does not know the distribution of problems,
 the space of solutions is not smooth, and the reward signal is
 uninformative, providing perhaps a few bits of information but not
 enough to steer the learner effectively. How can such a learner ever
 get off the ground? A common intuition is that if the solutions to
 these problems share a common structure, and the learner can solve
 some simple problems by brute force, it should be able to extract
 useful components from these solutions and, by composing them,
 explore the solution space more efficiently. Here, we formalize this
 intuition, where the solution space is the space of typed functional
 programs and the gained information is stored as a stochastic grammar
 over programs. We propose an iterative procedure for exploring such
 spaces: in the first step of each iteration, the learner explores a
 finite subset of the domain, guided by a stochastic grammar; in the
 second step, the learner compresses the successful solutions from the
 first step to estimate a new stochastic grammar. We test this
 procedure on symbolic regression and boolean circuit learning and
 show that the learner discovers modular concepts for these
 domains. Whereas the learner is able to solve almost none of the
 posed problems in the procedure's first iteration, it rapidly becomes
 able to solve a large number by gaining abstract knowledge of the
 structure of the solution space.

@@

Domain specific knowledge provides this structure. In local search, it
defines the similarity metric between solutions; in heuristic search
it provides the heuristic.

Here we ask how to bootstrap domain specific knowledge and extract
structure from the space of possible solutions by attempting to solve
multiple related problems in a given domain at once. Our approach is
motivated by the observation that many interesting problems -- such as
discovering scientific theories, developing new technologies, learning
to play a new instrument -- have solutions that are natrually
specified as programs.

@@

We can extend this representation with a simple polymorphic type
system~\cite{Pierce_2002}. In this representation, a type is either a
type primitive (e.g. reals, integers, booleans, etc.), a type variable
(e.g. $\sigma$), or a function type $ \tau_1 \rightarrow \tau_2$ of
functions from source type $\tau_1$ to target type $\tau_2$. Any combinator
can be represented as a binary tree~(Figure~\ref{fig:clbintree}) whose
leaves are typed primitive combinators and whose interior nodes
represent typed applications of combinators (we annotate interior
nodes with their types).

@@

A natural metric for the compressibility of a set of binary trees is
the number of unique subtrees (counting leaves) that this set
contains. More sophisticated metrics exist; we could use the negative
logarithm of the probability of the solutions under a suitable
probability distribution~\cite{DBLP:journals/tit/BarronRY98}. Future
work will explore whether this sophistication is worthwhile and which
prior on grammars makes sense. For our purposes here, the unique
subtree counting metric appropriately rewards reuse of useful
subtrees. As an example of this, suppose we have six tasks whose
solutions are integers $4^2, 5^2, 6^2, 7^2$, and $8^2$, and suppose
our frontier contains the $10$ expressions:
\vspace{-.3cm}
\captionsetup[subfigure]{labelformat=empty} 
\begin{figure}[h!]
   \begin{subfigure}{.17\linewidth}
     \begin{tikzpicture}[scale=.7, background rectangle/.style={fill=lightgray}] 
       \tikzset{level distance=15pt, sibling distance=10pt}
       \Tree [[  * 4 ] 4 ]
     \end{tikzpicture}
  \end{subfigure}
  \begin{subfigure}{.17\linewidth}
     \begin{tikzpicture}[scale=.7]
       \tikzset{level distance=15pt, sibling distance=10pt}
       \Tree [[  * 5 ] 5 ]
     \end{tikzpicture}
  \end{subfigure}
   \begin{subfigure}{.17\linewidth}
     \begin{tikzpicture} [scale=.7]
       \tikzset{level distance=15pt, sibling distance=10pt}
       \Tree [[  * 6 ] 6 ]
     \end{tikzpicture}
  \end{subfigure}
   \begin{subfigure}{.17\linewidth}
     \begin{tikzpicture}[scale=.7] 
       \tikzset{level distance=15pt, sibling distance=10pt}
       \Tree [[  * 7 ] 7 ]
     \end{tikzpicture}
  \end{subfigure}
  \begin{subfigure}{.17\linewidth}
     \begin{tikzpicture}[scale=.7] 
       \tikzset{level distance=15pt, sibling distance=10pt}
       \Tree [[  * 8 ] 8 ]
     \end{tikzpicture}
  \end{subfigure}
  \vspace{2mm}

@@

In the formulation we consider, each task $t$ corresponds to a
polynomial $f$, and an expression $e$ solves $t$ if it returns the
same value when applied $i \in 0, \dots, 9$ that $f$ does. In the
experiment shown here, the set of problems corresponds to the set of
all polynomials with degree less than or equal to 2. In the initial
library, we include the basic combinators I, S, B, C and four
arithmetic primitives, $1$, $0$, $*$, and $+$. The choice of initial
weights on these primitives can make a difference, as it determines
the relative ordering of combinators in the enumeration step. To get a
general sense of the algorithm's performance, we set the initial
weights to be slightly different on each run, perturbing them around a
uniform weighting.

@@

In Figure~\ref{fig:symregLearning}, we show performance results as we
vary the frontier size.Changing the frontier size changes the number
of tasks the algorithm identifies correctly over the course of 15
algorithm iterations (all experiments in this work were run for 15
iterations). As a baseline, we enumerated $10000*15=150000$
expressions from the initial grammar. This is the total number of
expressions that a run of the algorithm sees if it has a frontier size
of $10000$ and runs for $15$ iterations. Thus, if the benefit accruing
to the runs with larger frontier sizes is simply due to an increase in
the number of expressions seen, rather than increased learning, we
should see similarly good performance from this baseline run. In fact,
however, the baseline run only hits $27$ of the $1000$ tasks ($3\%$),
whereas our algorithm nears $100\%$ for a frontier of $10000$.

@@

As another demonstration of our approach, we investigate the
E.C. algorithm's ability to learn Boolean functions using only the logical
primitive NAND and the basic combinators. It is well known that a
boolean circuit can be constructed for any boolean function given only
the NAND gate. Here, the basic combinators take the place of the
wiring normally found in a circuit.

@@

We constructed two task sets. The first of these was constructed
explicitly to contain familiar modular structure. In this set of
tasks, we sampled $1000$ Boolean circuits using AND, OR, and NOT
gates. To accomplish this, we first sampled either 1, 2, 3 inputs,
then between 1 and 5 gates, randomly wiring the inputs of each new
gate to the output of one of the existing gates in the circuit. We
continued this sampling procedure until we had 1000 ``connected''
circuits, i.e., circuits all of whose outputs are wired to an input
except for the last one (the output gate). This resulted in $1000$
tasks consisting of $82$ unique Boolean functions, with a distribution
of tasks as shown in Figure~\ref{fig:booldistr}~(a). In
Figure~\ref{fig:boolLearningCurvesCircuits}, we present learning
curves for frontier sizes 100, 500 and 10000, using the same procedure
as in Section~\ref{sec:symreg}. In Figure~\ref{fig:booldistr}, we show
that the distribution over Boolean functions enumerated from the
grammar over expressions is much more similar to the true function
distribution after learning than before.

\begin{figure}
%%\vspace{-.3cm}
\includegraphics[width=\linewidth]{figures/circuitDistrComp2.png}
%%\vspace{-.6cm}
\caption{Distribution of boolean functions among (a) the set of 1000
  circuits randomly constructed from AND, OR and NOT gates (see text);
  b) the first 1000 expressions from the grammar over expressions
  \emph{before} learning; c) the first 1000 expressions from the
  grammar over expressions \emph{after} learning. Gray levels indicate
  percent of instances (circuits/expressions) which evaluate to that
  function.\label{fig:booldistr}}
%%\vspace{-.4cm}
\end{figure}

>>>>

\maketitle
\begin{abstract}
%\small
 Suppose a learner is faced with a domain of problems about which it
 knows nearly nothing. It does not know the distribution of problems,
 the space of solutions is not smooth, and the reward signal is
 uninformative, providing perhaps a few bits of information but not
 enough to steer the learner effectively. How can such a learner ever
 get off the ground? A common intuition is that if the solutions to
 these problems share a common structure, and the learner can solve
 some simple problems by brute force, it should be able to extract
 useful components from these solutions and, by composing them,
 explore the solution space more efficiently. Here, we formalize this
 intuition, where the solution space is the space of typed functional
 programs and the gained information is stored as a stochastic grammar
 over programs. We propose an iterative procedure for exploring such
 spaces: in the first step of each iteration, the learner explores a
 finite subset of the domain, guided by a stochastic grammar; in the
 second step, the learner compresses the successful solutions from the
 first step to estimate a new stochastic grammar. We test this
 procedure on symbolic regression and Boolean circuit learning and
 show that the learner discovers modular concepts for these
 domains. Whereas the learner is able to solve almost none of the
 posed problems in the procedure's first iteration, it rapidly becomes
 able to solve a large number by gaining abstract knowledge of the
 structure of the solution space.

@@

Domain specific knowledge provides this structure. In local search, it
defines the similarity metric between solutions; in heuristic search
it provides the heuristic. Here we ask how to bootstrap domain
specific knowledge and extract structure from the space of possible
solutions by attempting to solve multiple related problems in a given
domain at once. Our approach is motivated by the observation that many
interesting problems -- such as discovering scientific theories,
developing new technologies, learning to play a new instrument -- have
solutions that are natrually specified as programs.

@@

We can extend this representation with a simple polymorphic type
system~\cite{Pierce_2002}. In this representation, a type is either a
type primitive (e.g. reals, integers, Booleans, etc.), a type variable
(e.g. $\sigma$), or a function type $ \tau_1 \rightarrow \tau_2$ of
functions from source type $\tau_1$ to target type $\tau_2$. Any combinator
can be represented as a binary tree~(Figure~\ref{fig:clbintree}) whose
leaves are typed primitive combinators and whose interior nodes
represent typed applications of combinators (we annotate interior
nodes with their types).

@@

A natural metric for the compressibility of a set of binary trees is
the number of unique subtrees (counting leaves) that this set
contains. More sophisticated metrics exist; we could use the negative
logarithm of the probability of the solutions under a suitable
probability distribution~\cite{DBLP:journals/tit/BarronRY98}. Future
work will explore whether this sophistication is worthwhile and which
prior on grammars makes sense. For our purposes here, penalizing
according to the number of unique subtrees appropriately rewards reuse
of useful subtrees. As an example of this, suppose we have six tasks
whose solutions are integers $4^2, 5^2, 6^2, 7^2$, and $8^2$, and
suppose our frontier contains the $10$ expressions:
\vspace{-.3cm}
\captionsetup[subfigure]{labelformat=empty} 
\begin{figure}[h!]
   \begin{subfigure}{.17\linewidth}
     \begin{tikzpicture}[scale=.7, background rectangle/.style={fill=lightgray}] 
       \tikzset{level distance=15pt, sibling distance=10pt}
       \Tree [[  * 4 ] 4 ]
     \end{tikzpicture}
  \end{subfigure}
  \begin{subfigure}{.17\linewidth}
     \begin{tikzpicture}[scale=.7]
       \tikzset{level distance=15pt, sibling distance=10pt}
       \Tree [[  * 5 ] 5 ]
     \end{tikzpicture}
  \end{subfigure}
   \begin{subfigure}{.17\linewidth}
     \begin{tikzpicture} [scale=.7]
       \tikzset{level distance=15pt, sibling distance=10pt}
       \Tree [[  * 6 ] 6 ]
     \end{tikzpicture}
  \end{subfigure}
   \begin{subfigure}{.17\linewidth}
     \begin{tikzpicture}[scale=.7] 
       \tikzset{level distance=15pt, sibling distance=10pt}
       \Tree [[  * 7 ] 7 ]
     \end{tikzpicture}
  \end{subfigure}
  \begin{subfigure}{.17\linewidth}
     \begin{tikzpicture}[scale=.7] 
       \tikzset{level distance=15pt, sibling distance=10pt}
       \Tree [[  * 8 ] 8 ]
     \end{tikzpicture}
  \end{subfigure}
  \vspace{2mm}

@@

In the formulation we consider, each task $t$ corresponds to a
polynomial $f$, and an expression $e$ solves $t$ if it returns the
same value when applied to $i \in 0, \dots, 9$ that $f$ does. In the
experiment shown here, the set of problems corresponds to the set of
all polynomials with degree less than or equal to 2. In the initial
library, we include the basic combinators I, S, B, C and four
arithmetic primitives $1$, $0$, $*$, and $+$. The choice of initial
weights on these primitives can make a difference, as it determines
the relative ordering of combinators in the enumeration step. To get a
general sense of the algorithm's performance, we set the initial
weights to be slightly different on each run, perturbing them around a
uniform weighting.

@@

In Figure~\ref{fig:symregLearning}, we show performance results as we
vary the frontier size. Changing the frontier size changes the number
of tasks the algorithm identifies correctly over the course of 15
algorithm iterations (all experiments in this work were run for 15
iterations). As a baseline, we enumerated $10000*15=150000$
expressions from the initial grammar. This is the total number of
expressions that a run of the algorithm sees if it has a frontier size
of $10000$ and runs for $15$ iterations. Thus, if the benefit accruing
to the runs with larger frontier sizes is simply due to an increase in
the number of expressions seen, rather than increased learning, we
should see similarly good performance from this baseline run. In fact,
however, the baseline run only hits $27$ of the $1000$ tasks ($3\%$),
whereas our algorithm nears $100\%$ for a frontier of $10000$.

@@

As another demonstration of our approach, we investigate the
E.C. algorithm's ability to learn Boolean functions using only the logical
primitive NAND and the basic combinators. It is well known that a
Boolean circuit can be constructed for any Boolean function given only
the NAND gate. Here, the basic combinators take the place of the
wiring normally found in a circuit.

@@

We constructed two task sets. The first of these was constructed
explicitly to contain familiar modular structure. In this set of
tasks, we sampled $1000$ Boolean circuits using AND, OR, and NOT
gates. To accomplish this, we first sampled either 1, 2 or 3 inputs,
then between 1 and 5 gates, randomly wiring the inputs of each new
gate to the output of one of the existing gates in the circuit. We
continued this sampling procedure until we had 1000 ``connected''
circuits, i.e., circuits all of whose outputs are wired to an input
except for the last one (the output gate). This resulted in $1000$
tasks consisting of $82$ unique Boolean functions, with a distribution
of tasks as shown in Figure~\ref{fig:booldistr}(a). In
Figure~\ref{fig:boolLearningCurvesCircuits}, we present learning
curves for frontier sizes 100, 500 and 10000, using the same procedure
as in Section~\ref{sec:symreg}. In Figure~\ref{fig:booldistr}, we show
that the distribution over Boolean functions enumerated from the
grammar over expressions is much more similar to the true function
distribution after learning than before.

\begin{figure}
%%\vspace{-.3cm}
\includegraphics[width=\linewidth]{figures/circuitDistrComp2.png}
%%\vspace{-.6cm}
\caption{Distribution of Boolean functions among (a) the set of 1000
  circuits randomly constructed from AND, OR and NOT gates (see text);
  b) the first 1000 expressions from the grammar over expressions
  \emph{before} learning; c) the first 1000 expressions from the
  grammar over expressions \emph{after} learning. Gray levels indicate
  percent of instances (circuits/expressions) which evaluate to that
  function.\label{fig:booldistr}}
%%\vspace{-.4cm}
\end{figure}
